### 2024-02-12 parts/transducers.tex
big big rewrite

The goal of this section is to provide tools that do not require the assumption
of \kl{commutativity}. To that end, we introduce the notion of \kl{$k$-residual
transducer}, that generalizes the notion of residual transducer introduced for
\kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}. This notion of residual
transducer has its own interest, as it shifts the attention from (compatible)
equivalence relations of finite index that are the classical tool in automata
theory, towards (compatible) order relations that are \kl{well-quasi-ordered},
which is the order counterpart of having finite index.

While \kl{$k$-residual transducers} will provide canonical objects associated
to functions in $\NPoly$, their effective construction requires the ability to
decide $\NPoly$ inside $\ZPoly$, which is not currently known (except in the
commutative case). Furthermore, the correspondence between \kl{counters} in
this automaton and \kl{star-free $\Nat$-polyregular functions} is non-trivial,
and cannot be used \emph{as is} to decide membership in $\NSF$
based on this analysis of the \kl{$k$-residual transducer}.

\AP The upcoming sections will heavily refer to the notion of \intro{residual} of
a function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
over words in $\Sigma^*$.

\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
\reintro{good}. The notion of \reintro{good sequences} can also be applied to
binary relations that are not orderings, and a binary relation $R$ for which
every infinite sequence is \reintro{good} is said to be \intro{well}
\cite{MELL98}.

\subsection{Residual Transducers and WQOs}

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
transducer is based on the marble transducers \cite{EHB99}, and follows the
general pattern of transducers introduced in \cite{LOPEZ23b}.

@@

\begin{example}
    \label{non-canonical-transd:ex}
    Let $f \colon \set{a}^* \to \Nat$
    be defined as follows
    \begin{equation*}
        f(a^n) \defined
        \begin{cases}
            1 & \text{ if } n = 0 \\
            0 & \text{ if } n = 1 \\
            n & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}
    The function is computed by the two automata
    in \cref{non-canonical-transd:fig}.
\end{example}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.9\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.9\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    computing \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}

\AP As in the case of regular automata, we generalize the function $\delta
\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
$\lambda(q,\varepsilon) \defined 0$, and recursively define $\lambda(q, au)
\defined \app{\lambda(q,a)}{u} + \lambda(\delta(q,a), u)$. Following these
equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice
that if $\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q,
u) \in \mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.

\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
This notion can be applied to \kl{$\mathcal{H}$-transducers} by
considering counters of their underlying automata.

It is already known in the literature that this notion of transducers correctly
captures both $\NSF$ and $\NPoly$ in the sense that the following fact will
clarify.

\begin{fact}[Cite gaëtan]
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is computed by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{fact}

\AP Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if and only if $(g - f)
\in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if $(g - f) \in
\NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we are going to
study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f), \npolyleq[k])$. However,
to simplify notations, instead of writing $\app{f}{u} \npolyleq[k] \app{f}{v}$
when $u,v \in \Sigma^*$, we will use the convenient notation $u \resleq{f}{k}
v$, and directly consider the space $(\Sigma^*, \resleq{f}{k})$.

@@

Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
approach is to consider as states some representatives for the $\resleq{f}{k}$
minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
computing the same function (\cref{good-residual-ordering:fact}), and define
transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
\resleq{f}{k} ua$. This naïve approach does not yield a canonical model. For
instance, the function of \cref{non-canonical-transd:ex} can be computed in two
ways if we only consider $\resleq{f}{k}$ minimal elements, as witnessed in
\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
$\NSF[0]$, but one of the two presented transducers has a \kl{counter}. This is
why in \cref{residual-transducer:def}, we introduce the prefix ordering
$\prefleq$ to further constrain how the model computes.

\begin{definition}
    \label{residual-transducer:def}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{k-residual transducer}
    of $f$ 
    when
    it is a \kl{$\NPoly[k-1]$-transducer}
    satisfying the following properties:
    \begin{enumerate}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
            for $\prefleq$;
        \item $q_0 = \varepsilon$;
        \item every state $q \in Q$ is accessible from $q_0$;
        \item For all $u, a \in Q$,
            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
        \item For all $u,a \in Q$,
            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
    \end{enumerate}
\end{definition}

Because of the restriction on prefixes, the \kl{$k$-residual transducer}
is actually unique when it exists.

\begin{fact}
    \label{unique-res-transducer:fact}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    Then $f$ has at most one \kl{$k$-residual transducer}.
\end{fact}
\begin{proof}
    Let $\aTransd_1$ and $\aTransd_2$ be two
    \kl{$k$-residual transducers} for $f$.
    The two initial states must be $\varepsilon$.
    Let us prove by induction on $u \in \Sigma^*$ that
    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
    This will prove that 
    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.

    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
\end{proof}

We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
given a function $f$. Notice that this algorithm requires the ability to test
if a function belongs to $\NPoly[k]$, which is only known to be feasible for
\kl{commutative} polyregular functions. However, the termination of this
algorithm also proves the existence of the \kl{$k$-residual transducer}.

@@

\begin{fact}
    \label{q-o-prefix-cool:fact}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    At each step of the \texttt{while loop}
    of \cref{residual:algo}, the sets
    $Q$ and $O$ are such that
    \begin{enumerate}
        \item $Q \cup O$ is a \kl{downwards closed} subset of 
            $\Sigma^*$ for $\prefleq$;
        \item elements in $O$ are pairwise incomparable
            for $\prefleq$, and are maximal
            for $\prefleq$ inside $Q \cup O$.
    \end{enumerate}
\end{fact}

\begin{lemma}
    \label{correct-residual:lemma}
    If \cref{residual:algo} terminates on 
    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
    then it computes the \kl{$k$-residual transducer} of $f$.
\end{lemma}
\begin{proof}
    Because of \cref{q-o-prefix-cool:fact},
    we already know that $q_0 = \varepsilon$,
    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
    for $\prefleq$, 
    that every state of $Q$ is accessible from $q_0$.
    Notice that at every step,
    $\lambda(u,a)$ is defined as
    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
    at every step,
    we have that at step $i$,
    for all $ua \in O_i$,
    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
    which proves that the maximum considered in the algorithm
    is indeed computing correctly.
\end{proof}

@@

The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
\Rightarrow \resleq{f}{k})$ is \kl{well}, but $\resleq{f}{k}$ is a
\kl{well-quasi-ordering}. This has to be compared with the property of being
\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
increasing in every direction.

@@

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be \kl{combinatorial}.
    Then, $f$ is non-negative and for every pair
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
\end{conjecture}
\begin{proof}
    By induction on $k$.
\end{proof}

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be a non-negative function such that for every
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
    Then, $f$ belongs to $\NPoly[k]$.
\end{conjecture}
\begin{proof}
    By induction on $k$.
    For $k = 0$, non-negative is enough.
    To go beyond, we

\end{proof}

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ be a non-negative function, 
    and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is asumed to be \kl{commutative}, then all of the above
    properties are decidable, and the conversions effective.
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

\subsection{Aperiodicity and Counters}

\begin{lemma}
    \label{aperiodic-iff-residual:lem}
    Let $L$ be a regular language.
    The following are equivalent:
    \begin{enumerate}
        \item $L$ is aperiodic;
        \item The \kl{$0$-residual transducer} of $\ind{L}$
            is \kl{counter-free}.
    \end{enumerate}
\end{lemma}
\begin{proof}
    If the $0$-residual transducer is counter free,
    then $L$ is aperiodic because the minimal automata is also counter-free.

    % TODO
    \textbf{TODO Claim:}
    on every branch of the automata, the words have distinct residuals.

    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
    w_2) = t$.

    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
    strict prefix of $s_0$. It must exist, because otherwise the two are always
    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
    this contradicts the minimality of $u$.

    We have proven that $s_0$ and $s_1$ are comparable, hence they
    are equal.
    Finally, we have proven that $\delta(q, w_1) = s_0$,
    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.

    As a consequence $t$ and $q$ were comparable for the prefix relation,
    hence equal, and therefore $\delta(q, w) = q$.
\end{proof}

\begin{lemma}
    Let $\NSF[0] = \NPoly \cap \NSF[0]$,
    and elements of $\NSF[0]$
    have a \kl{counter-free} \kl{$0$-residual transducer}.
\end{lemma}
\begin{proof}
    TODO.
\end{proof}

\textbf{TODO:} one may think that we can ask that $u \resleq{f}{k} u w^n$
implies $u \resleq{f}{k} uw$, but it is false, as the following example will
demonstrate. This is why the definition of counter is more involved in our
case.

\begin{example}
    Let us consider $f(w) \defined (\card{w} - 3)^2$, 
    which is a \kl{star-free $\Nat$-polyregular function}.
    Then,
    $\app{f}{a} - f$ is not in $\NPoly$,
    but
    $\app{f}{a^10} - f$ is in $\NPoly$,
    hence there is a non-trivial counter.
\end{example}
\begin{proof}
    Remark that $\app{f}{a} = (\card{w} - 2)^2$,
    hence that $\app{f}{a}- f = 2\card{w} - 5$.
    Similarly, $\app{f}{a^10} - f = 20\card{w} + k$ where $k \geq 0$.

    Finally,
    let $n > 1$ be such that
    $\app{f}{a^n} - f$ belong to $\NPoly$.
    In this specific example,
    $\app{f}{a^n} = (\card{w} + n -3)^2$,
    and $\app{f}{a^n} - f = (2 \card{w} + n - 6)n$.
    The latter belongs to $\NPoly$ if and only if
    $n \geq 6$. Let $n \geq 6$, 
    remark that
    $\app{f}{a^n} - \app{f}{a^{n-1}}
    = 2 \card{w} + 2n - 7$. Because $n = n_0 + 6$ with $n_0 \geq 0$,
    we can rewrite the latter 
    as $2 \card{w} + 2n_0 + 5$, which belongs to $\NPoly$.
\end{proof}

\begin{lemma}
    \label{sf-no-periods-on-sequences:lemma}
    Let $f \colon \Sigma^* \to \Nat$ be a \kl{star-free $\Nat$-polyregular
    function}. Then,
    for all $u, w \in \Sigma^*$,
    there exists $n \in \Nat$ such that
    $u w^n \resleq{f}{k} u w^{n+1}$.
\end{lemma}
\begin{proof}[Proof sketch.]
    Let $t$ be a first order type,
    the map $g_t \colon X \mapsto \vcount{t}(uw^X)$ is a
    \kl{commutative} \kl{star-free $\Nat$-polyregular function},
    hence ultimately a polynomial, and therefore,
    translated enough, it is a $\Nat[X]$ polynomial.
    But this means that $g_t(X+1) - g_t(X) \in \Nat[X]$ for a large
    enough $X$, which is precisely what suffices.
    Now, we can do this for every first order type $t$
    with fewer variables and lower quantifier rank than those
    needed to define $f$, and conclude.
\end{proof}

Even though computability is not guaranteed, we have the following proof
scheme to decide aperiodicity of \kl{$\Nat$-polyregular functions}.
\begin{lemma}
    Let $f \in \NPoly[k]$ be \kl{ultimately polynomial},
    then every \kl{$k$-residual transducer} computing 
    $f$ has no counters, and has labels
    that are themselves \kl{ultimately polynomial}
    \kl{$\Nat$-polyregular functions}.
\end{lemma}
\begin{proof}

    We prove the result by induction on $k$. By definition
    of a residual transducer, the labels are \kl{ultimately polynomial}.
    Now, assume that there exists a counter
    in some \kl{$k$-residual transducer}
    $\aTransd \defined (Q, q_0, \delta, \lambda, F)$ of $f$. That is, there exists $\alpha \in \Sigma^*$,
    $w \in \Sigma^+$, and $n > 1$, such that
    $\delta(q_0,\alpha w^n) = \delta(q_0, \alpha)$.
    We want to prove that
    $\delta(q_0, \alpha w) = \delta(q_0, \alpha)$.
    Because $f$ is \kl{ultimately polynomial},
    it is in particular in $\ZSF[k]$, and therefore
    $g \defined \app{f}{\alpha w} - \app{f}{\alpha}$
    belongs to $\ZSF[k-1]$.
    It remains to be proven that this difference is a function in
    $\NPoly$.

    To that end, notice that the map $f(\alpha w^X \mathcal{M})$ is a
    \kl{commutative} \kl{star-free $\Nat$-polyregular function}. For a large
    enough $N_0$, we conclude that $f(\alpha w^{N_0+1} \mathcal{M}) - f(\alpha
    w^{N_0} \mathcal{M})$ belongs to $\NPoly$, hence that $\alpha w^{N_0}
    \resleq{f}{k-1} \alpha w^{N_0 + 1}$. Does that prove that $\alpha
    \resleq{f}{k-1} \alpha w$? Not at all and we cannot conclude.
\end{proof}

\begin{theorem}
    The following are equivalent for $f \in \NPoly[k]$.
    \begin{itemize}
        \item The function $f \in \NSF[k]$,
        \item $f$ is \kl{ultimately polynomial}.
        \item Every \kl{$k$-residual transducer} of $f$ has no counters,
        \item $f$ is \kl{computed} by some
            \kl{$\NSF[k-1]$-transducer}.
    \end{itemize}
\end{theorem}
\begin{proof}
    The only implication to prove is the first implies the last.
    We know that the automaton is counter free, so the languages of the states
    are star-free. Now, we just write 
    \begin{equation*}
        TODO
    \end{equation*}
    and we have finished the proof.
\end{proof}

Remark that the theorem is effective because being \kl{ultimately polynomial}
is an effective property! However, effective conversions are not possible as of
now, because the $k$-residual transducer algorithm uses an oracle to decide if
a function is in $\NPoly$.

\begin{lemma}
    \label{trivial-for-unary:lemma}
    Let $f \in \NPoly$ having unary input.
    In particular, $f$ is commutative.
    Then, $f \in \NSF$ if and only if 
    the \kl{residual transducer} of $f$
    is \kl{counter-free}.
\end{lemma}
\begin{proof}
    If there is a counter... well we will find it ^^
\end{proof}

\begin{conjecture}
    If $f$ is commutative, and the \kl{residual transducer}
    has a counter, then it has a single letter counter.
\end{conjecture}

\begin{conjecture}
    If $f$ is commutative, then 
    $f \in \NSF$ if and only if the \kl{residual transducer}
    of $f$ is \kl{counter-free}.
\end{conjecture}

\section{Trash polynomials}

\begin{definition}
    For every polynomial $P \in \Rel[X \vec{Y}]$, we define 
    the partial derivative with respect to $X$ as follows:
    \begin{equation*}
        \partial_X P \defined P(X + 1, \vec{Y}) - P(X, \vec{Y}) \quad .
    \end{equation*}
\end{definition}

\begin{fact}
    Let $P \in \CorrectPoly[X_1, \dots, X_n]$, then
    for all $1 \leq i \leq n$,
    $\partial_{X_i} P \in \CorrectPoly$.
\end{fact}
\begin{proof}
    We prove the result by induction on the number of variables.
    Let $\nu \colon \vec{X} \topartial \Nat$.
    Two cases, either $X_i \in \dom(\nu)$,
    in which case, we remark that
    we can split $\nu$ into two valuations,
    and that coefficients are good.

    Otherwise, evaluation commutes with partial,
    and we conclude by induction hypothesis.
\end{proof}

\begin{remark}
    The converse does not hold,
    for instance $P(X) \defined X - 1$ has all its derivatives
    that are in $\CorrectPoly$, but is not itself in $\CorrectPoly$.
\end{remark}

\begin{fact}
    Let $P \in \Nat[X,\vec{Y}]$, and $K,L \in \Nat$,
    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \Nat[X, \vec{Y}]$
    whenever $L \geq K$.
\end{fact}
\begin{proof}
    We can write
    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
    Now,
    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
    is precisely
    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$,
    remaining with positive coefficients.
\end{proof}

>>>>

\AP The goal of this section is to provide tools that do not require the
assumption of \kl{commutativity}. To that end, we introduce the notion of
\kl{$k$-residual transducer}, that generalizes the notion of residual
transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}.
This notion of residual transducer has its own interest, as it shifts the
attention from (compatible) equivalence relations of finite index that are the
classical tool in automata theory, towards (compatible) order relations that
are \kl{well-quasi-ordered}, which is the order counterpart of having finite
index. Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
\reintro{good}. The notion of \reintro{good sequences} can also be applied to
binary relations that are not orderings, and a binary relation $R$ for which
every infinite sequence is \reintro{good} is said to be \intro{well}
\cite{MELL98}.

The following definition of transducers is based on
the marble transducers \cite{EHB99}, and follows the general pattern of
transducers introduced in \cite{LOPEZ23b}.

@@

\begin{example}
    \label{non-canonical-transd:ex}
    Let $f \colon \set{a}^* \to \Nat$
    be defined as
    $f(\varepsilon) = 1$,
    $f(a) = 0$,
    and $f(a^n) = n - 1$ for all $n \geq 2$.
    \textbf{TODO: split the example}
    $f(\varepsilon) = 1$,
    $f(a) = 0$,
    $f(a^2) = 1$,
    and $f(a^n) = n - 3$ for all $n \geq 3$.
    The function is \kl{computed} by the two automata
    in \cref{non-canonical-transd:fig}.
\end{example}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    \kl{computing} the function of \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}

\AP As in the case of finite automata, we generalize the function $\delta
\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
$\lambda(q,\varepsilon) \defined 0$, and recursively define $\lambda(q, au)
\defined \app{\lambda(q,a)}{u} + \lambda(\delta(q,a), u)$. Following these
equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice that if
$\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q, u) \in
\mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.

\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
This notion can be applied to \kl{$\mathcal{H}$-transducers} by considering
counters of their underlying automata. It is already known in the literature
that this notion of transducers correctly captures both $\NSF$ and $\NPoly$ in
the sense that the following fact will clarify.

\begin{lemma}[{\cite{gaetanphd}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{lemma}

\AP The core concept of this section is that of a \intro{residual} of a
function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
over words in $\Sigma^*$. Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if
and only if $(g - f) \in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if
$(g - f) \in \NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we
are going to study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f),
\npolyleq[k])$. However, to simplify notations, instead of writing $\app{f}{u}
\npolyleq[k] \app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient
notation $u \resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
\resleq{f}{k})$.

@@

Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
approach is to consider as states some representatives for the $\resleq{f}{k}$
minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
computing the same function (\cref{good-residual-ordering:fact}), and define
transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
\resleq{f}{k} ua$. This naïve approach does not yield a canonical model which
is corrected by the following definition.

\begin{definition}
    \label{residual-transducer:def}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{$k$-residual transducer}
    of $f$ 
    when
    it is a \kl{$\NPoly[k-1]$-transducer}
    satisfying the following properties:
    \begin{enumerate}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
            for $\prefleq$;
        \item $q_0 = \varepsilon$;
        \item every state $q \in Q$ is accessible from $q_0$;
        \item For all $u, a \in Q$,
            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
        \item For all $u,a \in Q$,
            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
    \end{enumerate}
\end{definition}

Because of the restriction on prefixes, the \kl{$k$-residual transducer}
is actually unique when it exists.
We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
given a function $f$. Notice that this algorithm requires the ability to test
if a function belongs to $\NPoly[k]$, which is only known to be feasible for
\kl{commutative} polyregular functions. However, the termination of this
algorithm also proves the existence of the \kl{$k$-residual transducer}.

@@

\begin{lemma}
    \label{correct-residual:lemma}
    If \cref{residual:algo} terminates on 
    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
    then it computes the \kl{$k$-residual transducer} of $f$.
\end{lemma}

@@

The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quas-ordering $(\resleq{f}{k})$ is a
\kl{well-quasi-ordering}. This has to be compared with the property of being
\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
increasing in every direction.

@@

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ be a non-negative function, 
    and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item \label{n-poly-1-transd:item} $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item \label{n-poly-k:item} $f \in \NPoly[k]$;
        \item \label{n-poly-wqo:item} $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item \label{n-poly-well:item} every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item \label{n-poly-residual:item} $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is assumed to be \kl{commutative}, then all the above
    properties are decidable, and the conversions effective.
\end{theorem}
\begin{proof}
    \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
    \cref{transducer-nsf-npoly:lemma}. Then,
    \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
    \cref{n-poly-k-implies-wqo:lemma}.
    The implication \cref{n-poly-wqo:item} $\Rightarrow$ \cref{n-poly-well:item}
    is obvious.
    Then, \cref{wqo-implies-termination:lemma} proves
    that \cref{n-poly-well:item} implies \cref{n-poly-residual:item}.
    Finally, because a \kl{$k$-residual transducer} is a \kl{$\NPoly[k-1]$-transducer},
    \cref{n-poly-residual:item} implies \cref{n-poly-1-transd:item}.
\end{proof}
